Geometrical and physical interpretation of the action principle

### 2022-07-26 TeachingMaterial/ActionPrinciple/ActionPrincipleArticle.tex
ActionPrinciple: reviewed introduction

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics, which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. This gives a more physically motivated grounding of the classical theory and the proper setting to characterize both the physics and the geometry underlying the principle of stationary action. The assumption of determinism and reversible evolution corresponds to a diverge-free flow of the states, which preserves volumes in phase space (i.e. the count of states). The assumption of independence of the degrees of freedom means the volume reduces to the product of areas in each d.o.f., which is quantified by a closed two-form. The principle of stationary action arises as a feature of divergence-free fields and closed two-forms, as a general mathematical property of vector calculus and differential geometry\cite{souriau1997structure}. The additional assumption of equivalence between kinematics and dynamics (i.e. we can reconstruct the dynamical state simply by looking at the trajectory) is what allows us to express the principle in Lagrangian form. The argument can proceed in the reverse direction: assuming the principle of stationary action recovers a dynamical system that exhibit those three physical assumptions.

The mathematics needed to run the argument is not new, yet it is typically presented in a way and in a notation that is opaque to physicists. To provide a clear insight to the widest possible readership, we will first address the case of a single degree of freedom using standard vector calculus. This provides already all the geometrical and physical insights in a language that is accessible to all physicists and engineers. Then we proceed to the case of multiple independent degrees of freedom, which requires tools from differential geometry. Here we will use a notation that is similar to the one used in general relativity. We will concentrate mainly on the geometrical and physical meaning of the mathematical objects and their relationships, leaving the full argument and calculations in the appendix.

>>>>

As part of our larger project Assumptions of Physics, we developed an approach, called Reverse Physics, which examines current theories to find a set of starting physical assumptions that are sufficient to rederive them. We have found that Lagrangian mechanics is equivalent to three assumptions: determinism/reversibility, independence of degree of freedom and kinematic/dynamics equivalence. This gives a clear physically motivated understanding of the classical theory which allows us to characterize both the physics and the geometry underlying the principle of stationary action. What we find is that this arises as a general mathematical feature of divergence-free fields (and closed two-forms), which is the appropriate tool to describe a flow that conserves the number of states. The assumption of equivalence between kinematics and dynamics (i.e. we can reconstruct the dynamical state simply by looking at the trajectory) is what allows us to express the principle in the usual form. The argument can proceed in the reverse direction: assuming the principle of stationary action recovers a dynamical system that exhibit those three physical assumptions.

The mathematics needed to run the argument is not new\cite{souriau1997structure}. The two aspects that have been sorely lacking, and that we provide, are (1) a clear geometrical interpretation of the action principle and (2) a tight connection between the math and the physics it represents. To make the result accessible to the widest audience, we first cover the case of a single degree of freedom using standard vector calculus. We then proceed to the case of multiple independent degrees of freedom using tools from differential geometry. The main article will present all the key point needed to follow the argument and its physical and geometrical meaning, leaving the mathematical details and calculations to the appendix.
